---
title: "Quantitative Methods for HIV Researchers"
subtitle: "Dplyr"
author: "Janice M. McCarthy"
date: "November 5, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
setwd("~/2022-mic/computing/R")
```
# Manipulating data frames with `dplyr`

Once we have our data into a tidy format, we can use the functions in the `dplyr` packages to manipulate the rows and columns, get summaries or perform other operations on the data.

## Select

The `select` function extracts columns from the data frame. It can be used with explicit column names or 'helper' functions.

```{r}
starwars %>% head()

starwars %>%
select(name, birth_year, species) %>%
head(3)

starwars %>%
select(name, age=birth_year, species) %>%
head(3)

starwars %>% 
select(-name, -birth_year, -species) %>% 
head(3)

starwars%>% 
select(starts_with('s')) %>%
head(3)

starwars%>% 
select(matches("[aeiou]{2,}")) %>%
head(3)

starwars%>% 
select(name, age=birth_year, mass) %>%
arrange(age) %>%
head(3)

starwars%>% 
select(name, age=birth_year, mass) %>%
arrange(desc(age)) %>%
head(3)

starwars%>% 
select(name, age=birth_year, mass) %>%
arrange(age) %>%
head(5)

starwars%>% 
select(name, age=birth_year, mass) %>%
arrange(age, mass) %>%
head(5)

starwars%>% 
select(name, age=birth_year, mass) %>%
top_n(age, n=3)
```

### Exercises

Use the built-in data set `diamonds` to do the following:

1. Select only the `x, y` and `z` columns
2. Select all columns except `price`
3. Find the top 5 most expensive diamonds
4. Find the 5 most least expensive diamonds and print only their carat, cut, color and clarity
5. Find the 10 most expensive diamonds with the largest carat size
```{r}
diamonds %>% head()

diamonds %>%
select(x, y, z) %>%
head(3)

diamonds %>% 
select(-price) %>% 
head(3)

#starwars%>% 
#select(starts_with('s')) %>%
#head(3)

#starwars%>% 
#select(matches("[aeiou]{2,}")) %>%
#head(3)

diamonds%>% 
arrange(desc(price)) %>%
head(5)

diamonds%>% 
select(carat, cut, color, clarity, price) %>%
arrange(desc(price)) %>%
head(5)

#starwars%>% 
#select(name, age=birth_year, mass) %>%
#arrange(age) %>%
#head(5)

diamonds%>% 
arrange(desc(carat), desc(price)) %>%
head(5)

#starwars%>% 
#select(name, age=birth_year, mass) %>%
#top_n(age, n=3)
```
## Filter 

The `filter` function selects rows of a data frame according to a given criteria.

```{r}
starwars %>%
filter(birth_year > 100)

starwars %>%
filter((birth_year >= 600) | (name == 'Dooku')) # | stands for 'or' statement

starwars %>%
filter((birth_year >= 600) | (name == 'Dooku')) %>%
select(name, age=birth_year, homeworld)

starwars %>%
filter(birth_year > 100 , sex=='male') # , stands for and statement

starwars %>%
filter(str_detect(homeworld, "in")) %>% #look into the column homeworld and show string with "in" in them
select(name, homeworld) %>%
head(3)
```
### Exercises
Use the `diamonds` data set to do the following:

1. Find all diamonds whose cut is "Ideal"
2. Find all diamonds with carat > 2
3. Find all diamonds with carat > 2 and cut "Premiun" or "Ideal"
4. Find all diamonds with cut better than "Fair"
5. Use `filter` with `str_detect` to select all diamonds whose cut is 'Good' or 'Very Good'
```{r}
diamonds %>%
filter(cut == 'Ideal')

diamonds %>%
filter(carat > 2)

diamonds %>%
filter((carat > 2), (cut == 'Premium') | (cut == 'Ideal')) # | stands for 'or' statement

diamonds %>%
filter(cut != 'Fair')

diamonds %>%
filter(str_detect(cut, "Good"))


head(diamonds$cut) # tells you the ranking of any text categorical variables

diamonds %>% select(cut) %>% str()

```
## Mutate and transmutate

These functions perform operations across columns.
#mutate- adds onto dataframe; transmute- just make new columns and drop others

```{r}
starwars%>% 
select(name, age=birth_year, height, mass) %>%
head(3)

starwars%>% 
select(name, age=birth_year, height, mass) %>%
mutate(bmi=mass/(height/100)^2, obese=bmi>30) %>%
head(3)

starwars%>% 
select(name, age=birth_year, height, mass) %>%
transmute(bmi=mass/(height/100)^2, obese=bmi>30) %>%
head(3)

starwars %>%
transmute_if(is_character, str_to_upper) %>% # if the column is a character then make the string uppercase
head(3)

starwars %>%
select_at(c("birth_year", "height", "mass")) %>% #
head(3) 

starwars %>%
transmute_at(c("birth_year", "height", "mass"), ~ . + 1) %>%
head(3)
```
### Exercises

1. Add a new column to the `diamonds` data frame called "Class" that is 1 if the cut is "Ideal" or "Premium" **and** the carat is larger than 2, 0 if not.

```{r}
diamonds %>%
  ifelse(((cut == "Ideal" | cut == "Premium"), carat > 2), mutate (Class == 1), mutate (Class == 0)) %>%

head(10)
## Group by and summarize
```

```{r}
starwars %>%
summarize(avg_mass = mean(mass, na.rm=TRUE),
          median_age=median(birth_year, na.rm=TRUE)) # always a good idea to have the condition for removing entries containing 'na'

starwars %>%
summarize_if(is.numeric, mean, na.rm=TRUE)

starwars %>%
summarize_if(is.numeric, list(mean = mean, median = median), na.rm=TRUE)
 # mean = mean assigns the column name as 'mean'
starwars %>%
group_by(homeworld) %>%
summarize(avg_mass=mean(mass, na.rm=T)) %>%
view() # shows the whoel list: one of the entries is NA

starwars %>%
group_by(homeworld) %>%
summarize(avg_mass=mean(mass, na.rm=T)) %>%
filter(!is.na(homeworld)) %>%
head(3)

```

### Exercises

1. Group the diamonds by cut and find the mean in each category.
2. Count the number of diamonds of each color.

```{r}


diamonds %>%
group_by('cut') %>%
summarize_if(is.numeric, mean) %>%
head(10)

diamonds %>%
  group_by(color) %>%
  summarize(count = n())

```